Community-Conditioned Language Models

### 2021-05-14 paper/paper.tex
minor updates

\newcommand\Ppl{\mathsf{Ppl}}
\newcommand\IG{\mathsf{IG}}
\newcommand\Ind{\mathsf{Ind}}
\newcommand\LM{\mathrm{LM}}
\newcommand\CCLM{\mathrm{CCLM}}
\DeclareMathOperator*{\avg}{average}
\newcommand\BibTeX{B\textsc{ib}\TeX}

\newlist{hypotheses}{enumerate}{3}
\setlist[hypotheses,1]{parsep=0pt,itemsep=1pt,font=\bfseries,label=(H\arabic*)}

@@

In the present work, we investigate linguistic variation
across different online communities. In contrast to previous work, we
investigate variation across speech communities (rather than
macro-social categories), and use context-sensitive sequence models
capable of capturing variation beyond the lexical level.
%
We do so by investigating two questions.  First, we are interested in
variation in the level of linguistic complexity exhibited by different
communities.
%\jp{
  %Some examples (perhaps include if there is space):

@@

%TODO: outline of the paper.

@@

We investigate linguistic variation across various communities 
from the social media website Reddit.\footnote{Comments were obtained
  from the archive at \url{https://pushshift.io/}.
  \cite{Baumgartner2020}.}
%
Reddit is divided into forums called \textit{subreddits}, 
which are typically organized around a topic of interest. 
Users create \textit{posts}, which consist of a link, image, 
or text, along with a \emph{comment} section. 
Comments are threaded: a comment can be made directly on a post,
or appear as a reply to another comment.
%
Hereafter we refer to such comments as ``messages'', to match our
convention in mathematical formulas: the letter $c$ stands for a
community, and $m$ stands for a message.

Our dataset includes messages from \num{510} subreddits, 
sampled from the set of all forums 
with at least \num{5000} messages per month for each month
of the year 2015.
Each community corpus consist of \num{42000} randomly selected messages from the year 2015.
For each community, we reserve \num{1000} messages each for development and testing,
leaving a total of \num{20.4}M messages for training.

Messages were preprocessed as follows: 
We excluded the content of block quotes, code blocks, and tables.
We removed any markup (formatting) commands from the remaining text, 
extracting only rendered text.
We tokenized the messages using the default English model for the SpaCy tokenizer 
Version 2.2.3 \citep{Honnibal2017}.

@@

For the LSTM, we trained the models on a simple left-to-right language
modeling task with cross-entropy loss over the vocabulary.  Because
the transformer operates on all tokens in the sequence at once, the
language modeling was achieved by masking and incrementally un-masking
input tokens in a left-to-right fashion (so we have an
auto-regressive model).  We used the AdamW
\citep{Loshchilov2019} optimization algorithm, with an initial
learning rate of \num{0.001}, without any extra control on the decay
of learning rate.
%
We used a batch size of \num{256} and a maximum sequence length of
\num{64} tokens, truncating longer messages (16.8\% of messages were
longer than \num{64} tokens).  During training, a dropout rate of
$0.1$ was applied between encoder layers and after each linear layer.

@@

We trained the models until the validation loss stopped decreasing for
two epochs in a row, and used the weights from the epoch with the
smallest validation loss for testing.  In general, transformer models
trained for about half as long as the LSTM models (\cref{tab:model-results}, 
test epoch).

@@

We also consider the information gain of each CCLM over its baseline
un-conditioned counterpart with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_M(H_{\mathrm{LM}}(m))}}{e^{\avg_M(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m∈M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
Overall, the LSTM models perform better than the transformers, 
although the best transformer has higher information gain than the best LSTM.

@@

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best LSTM model . %TODO make long table
While the absolute performance ($\Ppl_{M_j}$) of the LSTM models is better,
we jhat the best transformer models have somewhat higher information gain
than their LSTM counterparts.

@@

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-info-gain}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are /r/counting, /r/friendsafari, and /r/Fireteams, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly strucutred and regular formats, which 
  are presumably conventional to these communities.}

@@

As an informal observation, we note that across all the models we
tested, communities where conditioning has the least effect tend to be
more organized around more general interest topics, such as the
\emph{relationships} and \emph{advice} subreddits, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organized around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialized vocabulary
and other linguistic patterns.

\section{Latent language model-based community classifiers}

The perplexity of a CCLM gives us a measure of linguistic complexity,
but it is not a measure of how linguistically \emph{specific} the
community is. However, the CCLM induces a latent classifier,
which we call a language model-based community classifier
(LMCC). Using the Bayes theorem, we can turn a
CCLM into a LMCC: the probability that a given message $m$
belongs to a community $c_i$ is calculated as follows:
\[P(c=c_i \mid m) = P(m | c_i)\frac {P(m)} {P(c=c_i)}\]
In the above,
$P(c=c_i)$ is the frequency of community $c_i$ in the dataset, and
$P(m \mid c_i)$ (the probability of a message $m$
in the context of community $c_i$) is estimated by the CCLM. 
$P(m)$ is the absolute
probability of a message, which can be computed as
\[P(m) = \sum_i P(c=c_i) P(m\mid c=c_i ). \]
Since the test set is balanced across communities,
we have an even prior for $P(c=c_i)$, giving us
\[P(c=c_i\mid m) \propto  P(m\mid c_i).\]

We can characterize the quality the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).

@@

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.96$ and $p < 0.0001$.

@@

@@

In this section we investigate whether (and how much) the community
embeddings obtained from the CCLMs correlate with the community-member
relationship on reddit.

To this end, we compare the CCLM-learned community
embedding with the community embedding created by \citet{Kumar2018},
who generated them using using a negative-sampling optimization
algorithm and the message-community co-occurrence matrix as ground
truth, using data which spans January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only depend on
whether a given user participates in a given
community. In sharp contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

@@

Computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). Each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

Further, we can apply a permutation to \(U\), \(V\) and \(Σ\) such that \(σ_i < σ_j\) iff \(i < j\). Doing so, the largest singular value \(σ_0\) corresponds to the principal
directions of correlation (\(U_0\), \(V_0\)), \(σ_1\) to the second
principal direction, etc.

@@

Thus if \(d(L,S)\) is below \(μ-3σ\), we can safely assume that there % below, because we have a distance rather than correlation in this convention.
is statistically significant correlation between \(L\) and
\(S\). However, below, we will also give a sense of \emph{how} the
correlation is manifested by analysis of the two principal components
of correlation in the linguistic embeddings, \(U_0\) and \(U_1\). To
do so we will plot the projection of embeddings on along \(U_0\) and
\(U_1\), together with the corresponding singular values.

(Another idea may be to correlate the cosine similarity between
\((S_i,S_j)\) and \((L_i,L_j)\). But this would yield one number for
each pair of communities \((510 × 509 / 2)\): too much data to deal with,
and it is hard to get a sense of the statistical significance of the
resulting correlation coefficient.)

>>>>

\newcommand\Ppl{\mathsf{Ppl}}
\newcommand\IG{\mathsf{IG}}
\newcommand\Ind{\mathsf{Ind}}
\newcommand\LM{\mathrm{LM}}
\newcommand\CCLM{\mathrm{CCLM}}
\DeclareMathOperator*{\avg}{average}
\newcommand\BibTeX{B\textsc{ib}\TeX}
\newcommand\subreddit[1]{\texttt{/r/#1}}
\newlist{hypotheses}{enumerate}{3}
\setlist[hypotheses,1]{parsep=0pt,itemsep=1pt,font=\bfseries,label=(H\arabic*)}

@@

In the present work, we investigate linguistic variation
across different online communities. In contrast to previous work, we
investigate variation across speech communities (rather than
macro-social categories), and use context-sensitive sequence models
capable of capturing variation beyond the lexical level.
%
We do so by investigating two questions.  First, we are interested in
variation in the level of linguistic complexity\jp{maybe revise terminology? Perhaps define linguistic complexity as perplexity?} exhibited by different
communities.
%\jp{
  %Some examples (perhaps include if there is space):

@@

\jp{outline of the paper is important here}

@@

We investigate linguistic variation across various communities 
from the social media website Reddit.\footnote{Comments were obtained
  from the archive at \url{https://pushshift.io/}.
  \cite{Baumgartner2020}.}
%
Reddit is divided into forums called \textit{subreddits}, 
which are typically organized around a topic of interest. 
Users create \textit{posts}, which consist of a link, image, 
or text, along with a \emph{comment} section. 
Comments are threaded: a comment can be made directly on a post,
or appear as a reply to another comment.
%
Hereafter we refer to such comments as ``messages'', to match our
convention in mathematical formulas: the letter $c$ stands for a
community, and $m$ stands for a message. \jp{what does \(p\) stand for?}

Our dataset includes messages from \num{510} subreddits, 
sampled\jp{what is the size of the whole population? why 510?} from the set of all forums 
with at least \num{5000} messages per month for each month
of the year 2015.
Each community corpus consist of \num{42000} randomly selected messages from the year 2015.
For each community, we reserve \num{1000} messages each for development and testing,
leaving a total of \num{20.4}M messages for training.

Messages were preprocessed as follows. 
We excluded the content of block quotes, code blocks, and tables.
We removed any markup (formatting) commands from the remaining text, 
extracting only rendered text.
We tokenized the messages using the default English model for the SpaCy tokenizer 
Version 2.2.3 \citep{Honnibal2017}.

@@

For the LSTM architecture, we trained the models on a simple left-to-right language
modeling task with cross-entropy loss.  Because
the transformer operates on all tokens in the sequence at once, the
language modeling was achieved by masking and incrementally un-masking
input tokens in a left-to-right fashion (so we have an
auto-regressive model).  We used the AdamW
\citep{Loshchilov2019} optimization algorithm, with an initial
learning rate of \num{0.001}, with no extra control on the decay
of learning rate.
%
We used a batch size of \num{256} and a maximum sequence length of
\num{64} tokens, truncating longer messages (16.8\% of messages were
longer than \num{64} tokens).  During training, a dropout rate of
$0.1$ was applied between encoder layers and after each linear layer.

@@

We trained the models until the validation loss stopped decreasing for
two epochs in a row, and used the weights from the epoch with the
smallest validation loss for testing.  In general, transformer models
were trained for about half as long as the LSTM models (\cref{tab:model-results}, 
test epoch).

@@

We also consider the average information gain per token of the CCLM over its baseline
un-conditioned counterpart, with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_M(H_{\mathrm{LM}}(m))}}{e^{\avg_M(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m∈M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
Overall, the LSTM models perform better than the transformers, 
although the best transformer has higher information gain than the best LSTM.

@@

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best LSTM model\jp{Is this the cclm or the baseline?}. %TODO make long table
While the absolute performance ($\Ppl_{M_j}$) of the LSTM models is better,
the best transformer models have somewhat higher information gain
than their LSTM counterparts.

@@

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-info-gain}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}

@@

As an informal observation, we note that across all the models we
tested, communities where conditioning has the least effect tend to be
more organized around more general interest topics, such as 
\subreddit{relationships} and \subreddit{advice}, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organized around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialized vocabulary
and other linguistic patterns.

\section{Latent language model-based community classifiers}
\jp{Perhaps use ``Latent Bayesian classifier'' instead.}
The perplexity of a CCLM gives us a measure of linguistic complexity\jp{terminology?},
but it is not a measure of how linguistically \emph{specific} the
community is. However, the CCLM induces a latent classifier,
which we call a language model-based community classifier
(LMCC).

So far, we have only loaded the community embedding which corresponds
to the actual community that the message originates from. By
evaluating the model's loss (and thus probability) for each potential community,
we can apply Bayesian reasoning to extract a latent community classifier for each message.
The probability that a given message $m$
belongs to a community $c_i$ is calculated as follows:
\[P(c=c_i \mid m) = P(m | c_i)\frac {P(m)} {P(c=c_i)}\]
In the above,
$P(c=c_i)$ is the frequency of community $c_i$ in the dataset, and
$P(m \mid c_i)$ (the probability of a message $m$
in the context of community $c_i$) is estimated by the CCLM. 
$P(m)$ is the absolute
probability of a message, which can be computed as
\[P(m) = \sum_i P(c=c_i) P(m\mid c=c_i ). \]
Since the test set is balanced across communities,
we have an even prior for $P(c=c_i)$, giving us
\[P(c=c_i\mid m) \propto  P(m\mid c_i).\]

We can characterize the quality\jp{performance} of the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).

@@

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.96$ and $Bp < 0.0001$.

@@

\jp{Conclusion: perplexity (linguistic complexity) is not correlated
  to how recognisable a community is. Presumably (in general) some very specific
  lexical items are characteristic of a community.}

@@

In this section we investigate whether (and how much) the community
embeddings obtained from the CCLMs correlate with the community-member
relationship on Reddit.

To this end, we compare the CCLM-learned community
embedding with the community embedding created by \citet{Kumar2018},
who generated them using using a negative-sampling optimization
algorithm and the message-community co-occurrence matrix as ground
truth, using data which spans January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only reflect the social
relationship between communities, via 
the user-community membership relation. In contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

@@

Computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

Further, we can apply a permutation to \(U\), \(V\) and \(Σ\) such
that \(σ_i > σ_j\) iff \(i < j\). Doing so, the largest singular value
\(σ_0\) corresponds to the principal directions of correlation
(\(U_0\), \(V_0\)), \(σ_1\) to the second principal direction, etc.

@@

Thus if \(d(L,S)\) is below \(μ_d-3σ_d\), we can safely assume that there % below, because we have a distance rather than correlation in this convention.
is statistically significant correlation between \(L\) and
\(S\). However, below, we will also give a sense of \emph{how} the
correlation is manifested by analysis of the two principal components
of correlation in the linguistic embeddings, \(U_0\) and \(U_1\). To
do so we will plot the projection of embeddings on along \(U_0\) and
\(U_1\), together with the corresponding singular values.

(Another idea may be to correlate the cosine similarity between
\((S_i,S_j)\) and \((L_i,L_j)\). But this would yield one number for
each pair of communities \((510 × 509 / 2)\): too much data to deal with,
and it is hard to get a sense of the meaning, let alone statistical significance of the
resulting correlation coefficient.)\jp{Not sure if this discussion has its place here.}
